pirouette: Determining the error BEAST2 makes in inferring a phylogeny

### 2019-03-07 article.tex
Use 'ht' float placement, as indicated by Overleaf

\verb;pirouette; can be installed easily from CRAN:
\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install.packages("pirouette")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
usethis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
library(pirouette)
\end{lstlisting}
Because \verb;pirouette; calls BEAST2, BEAST2 must be installed. 
This can be done from within R, using:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install_beast2()
\end{lstlisting}
For the option to select inference models,
\verb;pirouette; uses the "NS" BEAST2 package [\cite{maturana2018model}].
It can be installed from within R, using:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install_beast2_pkg("NS")
\end{lstlisting}

@@

We start from a tree simulated under a Yule (pure-birth) generative prior. 
We choose to use a tree with five taxa and a crown age of ten.

\begin{lstlisting}[
    language=R, floatplacement=H, frame=single, 
    label = {lst:create_yule_tree}, 
    caption = {Create a Yule tree}
  ]
phylogeny <- create_yule_tree(n_taxa = 5, crown_age = 10)
\end{lstlisting}

@@

\begin{lstlisting}[
    language=R,
    floatplacement=H, frame=single,
    label = {lst:create_alignment}, 
    caption = {Create an alignment}
  ]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=H, frame=single,
  label = {lst:create_generative_experiment},
  caption = {
    Create a default experiment, as described in Table~\ref{tbl:RQ1}.
  }
]
experiments <- list(create_experiment())
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
errors <- pir_run(
  phylogeny,
  pir_params = pir_params
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
pir_plot(errors)
\end{lstlisting}

@@

Here we start with a tree generated from an unknown 
diversification model, that has four taxa and a crown age of five:

\begin{lstlisting}[
  language=R, 
  floatplacement=H, 
  frame=single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model
]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
experiments <- create_all_experiments()
\end{lstlisting}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_2.png}
  \caption{
    Example 2: the error BEAST2 makes from a phylogeny when
    picking the best inference model.
    The x-axis specifies the tree type (which can be 'true' or 'twin'), the y-axis shows the error.
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_2}
\end{figure}

@@

Most of the settings for this example research question are
the same as in the previous section, which are
the phylogeny (listing \ref{lst:unknown_phylogeny}), 
the alignment parameters (listing \ref{lst:create_alignment}) 
and the list of experiments (listing \ref{lst:create_generative_experiment}).

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
twinning_params <- create_twinning_params()
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  twinning_params = twinning_params
)
\end{lstlisting}

@@

"What is the error made by BEAST2 from a phylogeny, when the generative model is known, compared to candidates with similar inference models, in relationship to the background error?"

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
generative_experiment <- create_experiment(
  model_type = "generative",
  run_if = "always",
  do_measure_evidence = TRUE
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
candidate_experiments <- create_all_experiments(
  site_models = list(create_jc69_site_model()),
  clock_models = list(create_strict_clock_model()),
  tree_priors = list(
    create_bd_tree_prior(), 
    create_ccp_tree_prior(), 
    create_cep_tree_prior()
  )
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
experiments <- c(list(generative_experiment), candidate_experiments)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  twinning_params = create_twinning_params()
)
\end{lstlisting}

>>>>

\verb;pirouette; can be installed easily from CRAN:
\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install.packages("pirouette")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
usethis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
library(pirouette)
\end{lstlisting}
Because \verb;pirouette; calls BEAST2, BEAST2 must be installed. 
This can be done from within R, using:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install_beast2()
\end{lstlisting}
For the option to select inference models,
\verb;pirouette; uses the "NS" BEAST2 package [\cite{maturana2018model}].
It can be installed from within R, using:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install_beast2_pkg("NS")
\end{lstlisting}

@@

We start from a tree simulated under a Yule (pure-birth) generative prior. 
We choose to use a tree with six taxa and a crown age of ten.

\begin{lstlisting}[
    language=R, floatplacement=ht, frame=single, 
    label = {lst:create_yule_tree}, 
    caption = {Create a Yule tree}
  ]
phylogeny <- create_yule_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}

@@

\begin{lstlisting}[
    language=R,
    floatplacement=ht, frame=single,
    label = {lst:create_alignment}, 
    caption = {Create an alignment}
  ]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, frame=single,
  label = {lst:create_generative_experiment},
  caption = {
    Create a default experiment, as described in Table~\ref{tbl:RQ1}.
  }
]
experiments <- list(create_experiment())
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
errors <- pir_run(
  phylogeny,
  pir_params = pir_params
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
pir_plot(errors)
\end{lstlisting}

@@

Here we start with a tree generated from an unknown 
diversification model, that has six taxa and a crown age of ten:

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
experiments <- create_all_experiments()
\end{lstlisting}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_2.png}
  \caption{
    Example 2: the error BEAST2 makes from a phylogeny when
    picking the best inference model.
    The x-axis specifies the tree type (which can be 'true' or 'twin'), 
    the y-axis shows the error
  }
  \label{fig:example_2}
\end{figure}

@@

Most of the settings for this example research question are
the same as in the previous section, which are
the phylogeny from an unknown speciation model (listing \ref{lst:unknown_phylogeny}), 
the alignment parameters (listing \ref{lst:create_alignment}) 
and the list of experiments (listing \ref{lst:create_generative_experiment}).

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
twinning_params <- create_twinning_params()
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  twinning_params = twinning_params
)
\end{lstlisting}

@@

"What is the error made by BEAST2 from a phylogeny, 
when the generative model is known, 
compared to candidates with similar inference models, 
in relationship to the background error?"
\richel{
  I think this is a nonsense research question and example:
  we should use a tree of unknown speciation model instead.
  Only then does twinning make sense
}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
generative_experiment <- create_experiment(
  model_type = "generative",
  run_if = "always",
  do_measure_evidence = TRUE
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
candidate_experiments <- create_all_experiments(
  site_models = list(create_jc69_site_model()),
  clock_models = list(create_strict_clock_model()),
  tree_priors = list(
    create_bd_tree_prior(), 
    create_ccp_tree_prior(), 
    create_cep_tree_prior()
  )
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
experiments <- c(list(generative_experiment), candidate_experiments)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  twinning_params = create_twinning_params()
)
\end{lstlisting}
